# Time, Space, Spacetime in R

Geostat-course 2015, Lancaster, summer school slides

\(
\newcommand{\E}{{\rm E}}       % E expectation operator
\newcommand{\Var}{{\rm Var}}   % Var variance operator
\newcommand{\Cov}{{\rm Cov}}   % Cov covariance operator
\newcommand{\Cor}{{\rm Corr}}
\)

## Data in R
Data in R are often organized in vectors,
```{r}
a = c(15, 20, 30)
a
b = c("blue", "brown", "blue")
b
```
in matrices
```{r}
m = matrix(1:4, 2, 2)
m
```
in `data.frame`s
```{r}
d = data.frame(a, b = c("blue", "brown", "brown"), c = c(65, 77, 69.6))
d
```
Such data has little meaning, because it is unclear what the variables
refer to, and what the records refer to. A more descriptive way would be
```{r}
Patients = data.frame(PatientID = a, EyeColor = b, Weight_kg = c(65, 77, 69.6))
Patients
```
where another table would be needed for the personal details related
to `PatientID`.

Note here that
 * records (rows in the `data.frame`) correspond to objects (persons), 
 * the variable `Weight_kg` has a [measurement unit](http://www.bipm.org/en/measurement-units/ "BIPM") encoded it its variable name

The weight numbers
```{r}
Patients$Weight_kg
```
carry no information about their measurement unit, allowing for meaningless computations such
as
```{r}
with(Patients, Weight_kg + PatientID)
```



## Temporal data in R

## Spatial data in R

### How far am I from home?
```{r}
options(geonamesUsername="edzer")
library(geonames)
pts = rbind(
 LA = GNsearch(name = "Lancaster University", adminCode1 = "ENG")[c("lng", "lat")],
 MS = GNsearch(name = "Münster")[1,c("lng", "lat")]
)
pts
```
It turns out that `pts` is still filled with character information, so we need to
transform it into numeric:
```{r}
class(pts$lng)
pts = sapply(pts, as.numeric)
rownames(pts) = c("LA", "MS")
pts
```
R has a function called `dist`, which computes (euclidian) distances in $\mathbb{R}^n$:
```{r}
dist(pts)
```

```{r}
library(sp)
pts = SpatialPoints(pts, CRS("+proj=longlat +datum=WGS84"))
```

```{r}
library(geosphere)
distHaversine(pts[1,],pts[2,])
distGeo(pts[1,],pts[2,])
spDists(pts)
```

## Spatiotemporal data in R

