---
title: "the Lancaster file"
output:
  html_document:
    toc: true
    theme: united
---

# Time, Space, Spacetime in R

Geostat-course 2015, Lancaster, summer school slides

\(
\newcommand{\E}{{\rm E}}       % E expectation operator
\newcommand{\Var}{{\rm Var}}   % Var variance operator
\newcommand{\Cov}{{\rm Cov}}   % Cov covariance operator
\newcommand{\Cor}{{\rm Corr}}
\)

## Data in R
Data in R are often organized in vectors,
```{r}
a = c(15, 20, 30)
a
b = c("blue", "brown", "blue")
b
```
in matrices
```{r}
m = matrix(1:4, 2, 2)
m
```
in `data.frame`s
```{r}
d = data.frame(a, b = c("blue", "brown", "brown"), c = c(65, 77, 69.6))
d
```
Such data has little meaning, because it is unclear what the variables
refer to, and what the records refer to. A more descriptive way would be
```{r}
Patients = data.frame(PatientID = a, EyeColor = b, Weight_kg = c(65, 77, 69.6))
Patients
```
where another table would be needed for the personal details related
to `PatientID`.

Note here that
 * records (rows in the `data.frame`) correspond to objects (persons), 
 * the variable `Weight_kg` has a [measurement unit](http://www.bipm.org/en/measurement-units/ "BIPM") encoded it its variable name

The weight numbers
```{r}
Patients$Weight_kg
```
carry no information about their measurement unit, allowing for meaningless computations such as
```{r}
with(Patients, Weight_kg + PatientID)
```
The answer is correct, though, as `+` requires two numeric arguments. If we try to add
eye color to body weight, we get a warning if `EyeColor` is a `factor` (by default,
`data.frame` converts `character` vectors into `factor`s!),
```{r}
with(Patients, EyeColor + Weight_kg)
```
or in case `EyeColor` is `character`, we get an error:
```{r}
Patients$EyeColor = as.character(Patients$EyeColor)
with(Patients, EyeColor + Weight_kg)
```
Take home messages:

1. `data.frame` may change your information (character to `factor`)
1. computations that are syntactically correct are not necessarily meaningful
1. *never ever* ignore errors or warnings, but drill down into where they come from

Exercises

1. how can the `data.frame` command above be modified such that
the character variable is not modified into a `factor`?
1. how can R be configured such that this is always the case?
1. in a long script, warnings are printed at the end. How can
you make them print where they happen, or convert them into
errors? (hint: `?options`)

## Temporal data in R

### How do we represent time?
[ISO 8601](https://en.wikipedia.org/wiki/ISO_8601) -
Data elements and interchange formats – Information interchange – Representation of dates and times

### Who knows what time it is? 
Coordinated Universal Time
[UTC](https://en.wikipedia.org/wiki/International_Organization_for_Standardization "UTC")
``It is, within about 1 second, mean solar time at 0 degree
longitude; it does not observe daylight saving time.'' (formerly: GMT). It keeps track of
the Earth's rotation and International Atomic Time ([TAI](https://en.wikipedia.org/wiki/International_Atomic_Time)), by introducing leap seconds [now
and then](https://en.wikipedia.org/wiki/Leap_second).
Real-time approximations to UTC are broadcasted wireless by GPS and
radio time signals; computers usually use the [network time
protocol](https://en.wikipedia.org/wiki/Network_Time_Protocol "NTP")
to synchronize clocks.

### Does R know all this?

Yes! To find out what time it is, R of
course relies on the system clock. The R [source
code](https://stat.ethz.ch/pipermail/r-devel/2015-July/071472.html)
is updated every time a leap second is announced. It is based on
the code that operating systems use to handle date/time.

## Spatial data in R

### How far am I from home?
[Geonames](http://www.geonames.org/) is a gazetteer
(geographical dictionary or directory) that has a collection
of place names, place types and geographical coordinates in
[WGS84](https://en.wikipedia.org/wiki/World_Geodetic_System),
along with several other things.

You can register yourself at [geonames](http://www.geonames.org/login).
```{r eval=FALSE}
options(geonamesUsername = "edzer") # don't use my name!
library(geonames)
pts = rbind(
 LA = GNsearch(name = "Lancaster University", adminCode1 = "ENG"),
 MS = GNsearch(name = "Münster")[1,]
)[c("lng", "lat")]
```
```{r echo=FALSE}
load("pts.RData")
```
```{r}
pts
```

It turns out that `pts` is still filled with character information, 
```{r}
class(pts$lng)
```
so we need to transform it into numeric:
```{r}
pts = sapply(pts, as.numeric) # this drops rownames, so:
rownames(pts) = c("LA", "MS") # Lancaster, Muenster
pts
```
R has a function called `dist`, which computes (euclidian) distances in $\mathbb{R}^n$:
```{r}
dist(pts)
```
which returns `10.6`, but 10.6 what? Degrees? Nautical miles? Clearly, our data are
not in planar space, so we need to be more clever. Package `sp` 
```{r}
library(sp)
pts = SpatialPoints(pts, CRS("+proj=longlat +datum=WGS84"))
spDists(pts)
```

```{r}
library(geosphere)
distHaversine(pts[1,],pts[2,])
distGeo(pts[1,],pts[2,])
```

## Spatiotemporal data in R

