---
title: "Geostat 2015, Lancaster geostat-course.org slides"
output:
  html_document:
    toc: true
    theme: united
---

[CC-BY-SA](http://creativecommons.org/licenses/by-sa/4.0/), Edzer Pebesma 2015.

# Time, Space, Spacetime in R

## Data in R
Data in R are often organized in vectors,
```{r}
a = c(15, 20, 30)
a
b = c("blue", "brown", "blue")
b
```
in matrices
```{r}
m = matrix(1:4, 2, 2)
m
```
or in `data.frame`s
```{r}
d = data.frame(a, b = c("blue", "brown", "brown"), c = c(65, 77, 69.6))
d
```
Such data has little meaning, because it is unclear what the variables
refer to, and what the records refer to. A more descriptive way would be
```{r}
Patients = data.frame(PatientID = a, EyeColor = b, Weight_kg = c(65, 77, 69.6))
Patients
```
where another table would be needed for the personal details related
to `PatientID`.

Note here that
 * records (rows in the `data.frame`) correspond to objects (persons), 
 * the variable `Weight_kg` has a [measurement unit](http://www.bipm.org/en/measurement-units/ "BIPM") encoded it its variable name

The weight numbers
```{r}
Patients$Weight_kg
```
carry no information about their measurement unit, allowing for meaningless computations such as
```{r}
with(Patients, Weight_kg + PatientID)
```
The answer is correct, though, as `+` requires two numeric arguments. If we try to add
eye color to body weight, we get a warning if `EyeColor` is a `factor` (by default,
`data.frame` converts `character` vectors into `factor`s!),
```{r}
with(Patients, EyeColor + Weight_kg)
```
or in case `EyeColor` is `character`, we get an error:
```{r}
Patients$EyeColor = as.character(Patients$EyeColor)
print(try(with(Patients, EyeColor + Weight_kg)))
```
Take home messages:

1. `data.frame` may change your information (convert character into `factor`), which
is useful for some purposes (categorical variables, models), but not for others (DNA sequences).
1. computations that are syntactically correct are not necessarily meaningful
1. *never ever* ignore errors or warnings, but drill down into where they come from
1. Records in a table (rows in a `data.frame`) often refer to *objects*, variables (columns) often to properties of these objects.

Exercises

1. how can the `data.frame` command above be modified such that
the character variable is not modified into a `factor`?
2. how can R be configured such that this is always the case?
3. in a long script, warnings are printed at the end. How can
you make them print where they happen, or convert them into
errors? (hint: `?options`)
4. how is a `data.frame` related to a `list`?
5. try to understand the following commands
 * `d[1,]`
 * `d[,1]`
 * `d[,1,drop=FALSE]`
 * `d[1,1]`
 * `d[1,1,drop=FALSE]`
 * `d[1]`
 * `d["a"]`
 * `d[["a"]]`
 * `d$a`
 * `d[,"a"]`
 * `d[,c("a","b")]`
 as you will see, the logic behind selection for spatial, temporal, and spatiotemporal objects follows this very closely. 
6. `matrix` and `data.frame` both represent two-dimensional organisations of data; why don't we use `matrix` for everything?

## Temporal data in R

### How do we represent time?

People communicate time by using words, which can be written down as character
information. Phrases like *Sunday, 16 Aug 2015* and *18:36* are widely understood
(although language dependent) to indicate date and time. Combined, we sometime see
```{r}
system("date > date.file") # works on unix systems!
readLines("date.file")     # read & print back in R
```
There are actually a lot of different ways in which we can represent
time information, and still understand it. For computers, this is less
easy:
```{r}
data(wind, package = "gstat")
head(wind[,1:7])
data(Produc, package = "plm")
head(Produc[,1:6])
library(foreign)
read.dbf(system.file("shapes/sids.dbf", package="maptools"))[1:5,c(5,9:14)]
```

### ISO 8601

[ISO 8601](https://en.wikipedia.org/wiki/ISO_8601) - <em>Data elements
and interchange formats - Information interchange - Representation
of dates and times</em> tries to create some order in this mess. It uses
e.g. `2015-08-07` for Aug 7 2015, and `2015-08-07T18:30:27+00:00`
to specify a given time in a particular time zone on that date.

ISO 8601 implicitly defines time stamps to refer to
time *intervals*: 2015-08-07 refers to the *full day* of Aug 7, i.e.
from 2015-08-07 00:00 up to but not including 2015-08-08 00:00. The
time stamp 2015-08-07 00:00 has a duration of one second. 2015-08
refers to a full month.

### Who knows what time it is? 

Coordinated Universal Time
[UTC](https://en.wikipedia.org/wiki/International_Organization_for_Standardization "UTC")
``It is, within about 1 second, mean solar time at 0 degree
longitude; it does not observe daylight saving time.'' (formerly: GMT). It keeps track of
the Earth's rotation and International Atomic Time ([TAI](https://en.wikipedia.org/wiki/International_Atomic_Time)), by introducing leap seconds [now and then](https://en.wikipedia.org/wiki/Leap_second). Real-time approximations to UTC are broadcast wireless by GPS and radio time signals; computers usually use the [network time protocol](https://en.wikipedia.org/wiki/Network_Time_Protocol "NTP") to synchronize clocks.

### Does R know all this?

Kind of - for instance 
```{r}
as.Date(.leap.seconds)
```
To find out what time it is, R of course relies on the system clock. The R [source code](https://stat.ethz.ch/pipermail/r-devel/2015-July/071472.html) is updated every time a new leap second is announced. It is based on the code that operating systems use to handle date/time.

### Representing time in R

Although one could argue that time is represented by a real number ($\mathbb{R}$), for the sequence
```{r}
c(55, 60, 65, 70)
```
it is not clear to which points in time we refer to: are these years since 1900, or seconds since the start of Today? To fix this, one needs to set an offset (when is zero?) and a unit (how long takes a unit increas?).

R has two built-in formats: for date it has `Date`, for time `POSIXt`:
```{r}
(d = Sys.Date())
class(d)
print(as.numeric(d), digits = 15)
(t = Sys.time())
class(t)
print(as.numeric(t), digits = 15)
```
`Date` is stored as the (integer) number of days since 1970-01-01 (in the local time zone), time is stored as the (fractional) number of seconds since 1970-01-01 00:00 UTC. We can see that time is *printed* in the current time zone.

R also ``understands'' time differences, e.g.
```{r}
d - (d+1)
d - (d+1.1)
t - (t+24*3600)
t - t+24*3600
```

Exercise

1. why do the last two expressions print a different result?

`POSIXct` uses one double to representation a time instance, `POSIXlt` represents time as a list (each time stamp one list), and is convenient to isolate time
components:
```{r}
t
(lt = as.POSIXlt(t))
unlist(lt)
sapply(lt, class)
lt$sec
```

Exercise

1. explain all the fields of a POSIXlt object

### Time zones in R

Time zones make it easy to understand whether it is day or night
without asking where you are, and take care of daylight saving time.
There is a [time zone database](https://www.iana.org/time-zones)
and a [wikipedia entry](https://en.wikipedia.org/wiki/Tz_database),
and R uses this either directly or through the GNU C library.

```{r}
Sys.setenv(TZ="CET")
as.POSIXct("2015-03-28") - as.POSIXct("2015-03-27") # regular
as.POSIXct("2015-03-30") - as.POSIXct("2015-03-29") # 23 hrs: DST starts
as.POSIXct("2015-10-26") - as.POSIXct("2015-10-25") # 25 hours: DST ends
```
whereas in UTC,
```{r}
Sys.setenv(TZ="UTC")
as.POSIXct("2015-03-30") - as.POSIXct("2015-03-29")
as.POSIXct("2015-10-26") - as.POSIXct("2015-10-25")
Sys.setenv(TZ="CET")
```
having irregular intervals (day lengths) makes it complicated
to compute daily means from e.g. hourly values; working with UTC
may solve this problem. Climate scientists often model climate
for years with 360 days, to get rid of the complexity of handling
unequal month lengths and leap years. Aligning such data with POSIX
time would be fun!

### Reading time data into R

Date and time are understood from character strings (or factor)
when entered like this:
```{r}
as.Date("2015-08-17")
as.POSIXct("2015-08-17")
as.POSIXct("2015-08-17 12:25")
```
but not like this
```{r}
as.POSIXct("2015-08-07T18:30:27Z")
```
where everything after the T is ignored, without warning! 

Arbitrary time formats can be specified using `strptime`, e.g.
```{r}
strptime("2015-08-07T18:30:27Z", "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
```

The (implicit) interval information of ISO 8601 time expressions
is lost when read as POSIXt objects, as
```{r}
identical(as.POSIXct("2015-08-17"), as.POSIXct("2015-08-17 00:00"))
```

To convert numerical values into dates or time, the offset needs to
be specified:
```{r}
print(try(as.Date(365)))
as.Date(365, origin = "1970-01-01")
as.POSIXct(365 * 24 * 3600, origin = "1970-01-01")
```
although package zoo takes a somewhat different take on this:
```{r}
library(zoo)
as.Date(365)
```


### Packages with tools to deal with time

The CRAN [Time Series Analysis Task View](https://cran.uni-muenster.de/web/views/TimeSeries.html) describes all packages dealing with time, time series data, and its analysis.
Several packages provide other ways of representing time; package [lubridate](https://cran.uni-muenster.de/web/packages/lubridate/index.html) for instance provides explicit representation of time intervals.

### Packages for analyzing time series data

The time series task view is complete here, but I will mention two important packages: [zoo](https://cran.uni-muenster.de/package=zoo) and [xts](https://cran.uni-muenster.de/package=xts).

Zoo is for ordered observations, where the ordering does not *have* to be time:
```{r}
library(zoo)
zoo(rnorm(10), 1:10)
```
it provides a lot of convenient functions, including `aggregate`, `na.fill`, and e.g. moving average filters. See for instance `?aggregate.zoo` for many examples aggregating time series to daily, monthly, quarterly or yearly values.

Package xts builds on top of zoo (xts is a subclass of zoo), and requires that data are ordered by time. It allows for different time classes (e.g. `Date` and `POSIXct`) but works with `POSIXct` under the hood. xts adds very convenient ISO 8601 style time interval selection:
```{r}
library(xts)
x = xts(1:365, as.Date("2015-01-01")+0:364)
nrow(x)
nrow(x["2015-05"]) # all days in May
nrow(x["2015-05/06"]) # all days in May and June
```

## Spatial data in R

### How far from home am I?
[Geonames](http://www.geonames.org/) is a gazetteer
(geographical dictionary or directory) that has a collection
of place names, place types and geographical coordinates in
[WGS84](https://en.wikipedia.org/wiki/World_Geodetic_System),
along with several other things.

You can register yourself at [geonames](http://www.geonames.org/login).
```{r eval=FALSE}
options(geonamesUsername = "myUserName") # register, then use your own name
library(geonames)
pts = rbind(
 LA = GNsearch(name = "Lancaster University", adminCode1 = "ENG"),
 MS = GNsearch(name = "Münster")[1,]
)[c("lng", "lat")]
```
```{r echo=FALSE}
load("pts.RData")
```
```{r}
pts
```

It turns out that `pts` is still filled with character information, 
```{r}
class(pts$lng)
```
so we need to transform it into numeric:
```{r}
pts = sapply(pts, as.numeric) # this drops rownames, so:
rownames(pts) = c("LA", "MS") # Lancaster, Muenster
pts
```
R has a function called `dist`, which computes (euclidean) distances in $\mathbb{R}^n$:
```{r}
dist(pts)
```
which returns `10.6`, but 10.6 what? Degrees? Nautical miles? Oranges? No,
plane nonsense, and `dist` is only useful for geographical coordinates (long/lat) in small areas close to the equator: our data are not in planar space, so we need to be more clever. 

Package `sp` provides classes for points, lines, polygons and grids which register their coordinate reference system (CRS), and chooses appropriate distance functions:
```{r}
library(sp)
pts = SpatialPoints(pts, CRS("+proj=longlat +datum=WGS84"))
spDists(pts) # km, great circle distance over the WGS84 ellipsoid
```
package [geosphere](https://cran.uni-muenster.de/package=geosphere) provides several alternative spherical/ellipsoidal distance measures:
```{r}
library(geosphere)
distHaversine(pts[1],pts[2])
distGeo(pts[1],pts[2]) # different from spDist:
spDists(pts)[2,1]*1000 - distGeo(pts[1],pts[2])  # difference in m
```

### Subsetting sp objects
as we have seen, a `SpatialPoints` object can be subsetted by its index, but
also by its row name; `pts[1]` and `pts[1,]` yield the same: we think of these geometry-only objects as having only records:
```{r}
pts[1] # selects record 1, but looks ambiguous
pts[1,] # looks less ambiguous
pts["MS",]
```

If we add attributes to `pts`, e.g. by
```{r}
pts$pop = c(299.708, 45.952) # according to Wikipedia
pts$area = c(302.9, 19.2) # I asked google
pts
```
we see that the class of `pts` now has changed from `SpatialPoints` to `SpatialPointsDataFrame`, the reason being that in addition to geometry (locations), the object now has attributes (in a `data.frame` slot). We can subset or manipulate these attributes in the same fashion as we can with plane `data.frame`s:
```{r}
pts[1] # now selects a variable!
pts[,1] # does the same, less ambiguous
pts["pop"] # does the same, by name
pts[,"pop"] # does the same
pts[["pop"]] # extract vector
pts$pop      # synonymous
pts[1,"pop"] # select 1 geometry, 1 variable
pts[1,]$pop  # select 1 geometry, extract variale
(pts$popDensity = pts$pop / pts$area) # create & add new variable
```

We can even select using the spatial predicate *intersect*, as in
```{r}
library(spacetime)
data(air) # loads the boundary of Germany, called DE
proj4string(DE) = proj4string(pts) # semantically identical, but syntactically different
pts[DE,] # selects the point(s) in pts intersecting with polygon DE
plot(DE, axes = TRUE)
points(pts[DE,], col = 'red', pch = 16)
```

Note that

1. Coordinate reference systems (CRS) need to be defined in order to decide which distance metrics are meaningful, and whether objects can be meaningfully combined by comparing coordinates (e.g., `pts` and `DE`)
1. `sp` objects register coordinate reference systems, and warn/err in case of  mismatch

Exercises

1. Create a login for geonames, and retrieve the coordinates of your home address, or else use some other gazetteer to retrieve those
1. Compute how far away from home you currently are, using different distance metrics.

## Spatiotemporal data in R

The [SpatioTempofal Task View](https://cran.uni-muenster.de/web/views/SpatioTemporal.html) describes the packages that represent and/or analyze spatiotemporal data available on CRAN.

### How are spatiotemporal data organized in package spacetime?

Essentially, spacetime distinguishes between three types of spatiotemporal data: 

1. regular data where for fixed locations (or regions) observations are collected for the same time instances or intervals (e.g. socio-economic data or data from fixed sensors)
2. irregular data for which observations are arbitrarily distributed in space-time
3. trajectories, data for moving objects.

Support for trajectories in package `spacetime` is rudimentary, and is more developed in package `trajectories` (see below)

Irregular spacetime data are stored in `STI` objects, or `STIDF` if they have attributes (magnitude of an earth quake, cause of a disease):
```{r}
showClass("STI")
```
the `sp` and `time` slots need to have the same number of records, and are simply matched by order to identify the location and time of a particular event. A
`data` slot of `STIDF` has the same number of records and is matched accordingly.

Regular spacetime data have recurrent observations for fixed spatial entities. The `STF`, space-time-full, data structure represents this by
```{r}
showClass("STF")
```
Although the class structure seems identical, `sp` and `time` may have different number of records, and the assumption here is that every geometry record has a time series of data of length `nrow(time)`. The `data` slot of `STFDF` has `length(sp) * nrow(time)` records, space cycling fastest.

A third type,  `STF` (space-time-sparse), defined as
```{r}
showClass("STS")
```
mimics `STF` but does not store all cell (space x time) values; it keeps an index vector to those cells that are filled, in order to be efficient for very sparse ST regular layouts.

Note that

1. the `sp` slot can contain any type of spatial data (points, lines, polygons, grid)
2. time does not need to be regular, only the sequence of time points (or intervals) is identical for each space feature
3. `endTime` can be used to define intervals; by default `STI` objects have zero interval width (indicating time instance), `STF` have interval length equal to the time to next observation (consecutive intervals).

Quite often, regular or irregular data are aggregated to new spatial and/or temporal units. Package `spacetime` tries hard to accomodate all possibilities.

### Analysis of regular spatiotemporal data (air quality sensor data)

Package `spacetime` contains a dataset called `air`, which contains
daily PM$_{10}$ values for rural air quality stations over Germany,
from 1998-2009.
```{r}
library(sp)
library(spacetime)
data(air)
rural = STFDF(stations, dates, data.frame(PM10 = as.vector(air)))
dim(rural)
stbox(rural)
```

We now aggregate the 2008 daily values to NUTS-1 region-average daily values by
```{r}
x = as(rural[,"2008"], "xts")
apply(x, 1, mean, na.rm=TRUE)[1:5]
dim(rural[,"2008"])
x = aggregate(rural[,"2008"], DE_NUTS1, mean, na.rm=TRUE)
dim(x)
stplot(x, mode = "tp", par.strip.text = list(cex=.6))
```

We can aggregate to the complete region, by which the object becomes a `xts` object (unless we would specify `simplify=FALSE` in the call to `aggregate`):
```{r}
x = aggregate(rural[,"2008"], DE, mean, na.rm=TRUE)
class(x)
plot(x[,"PM10"])
```

Some stations contain only missing values for certain time periods; we 
can de-select those, and aggregate to monthly means:
```{r}
x = as(rural[,"2008"], "xts")
apply(x, 2, mean, na.rm=TRUE)[1:5]
sel = which(!apply(as(rural[,"2008"], "xts"), 2, function(x) all(is.na(x))))
x = aggregate(rural[sel, "2008"], "month", mean, na.rm=TRUE)
stplot(x, mode = "tp", par.strip.text = list(cex=.6))
```

Next, we use `zoo::as.yearqtr` to aggregate to quarterly values:
```{r}
library(zoo)
x = aggregate(rural[sel,"2005::2011"], as.yearqtr, median, na.rm=TRUE)
stplot(x, mode = "tp", par.strip.text = list(cex=.6))
```

Finally, we compute a yearly mean values for 2008 and 2009, for the whole
country:
```{r}
DE.years = STF(DE, as.Date(c("2008-01-01", "2009-01-01")))
aggregate(rural[,"2008::2009"], DE.years, mean, na.rm=TRUE)
```

We can do some space-time geostatistics on these data, e.g. by

```{r}
rr = rural[,"2005::2010"]
unsel = which(apply(as(rr, "xts"), 2, function(x) all(is.na(x))))
r5to10 = rr[-unsel,] # remove series that are empty for this time period
summary(r5to10)
dim(r5to10)
rn = row.names(r5to10@sp)[4:7]
rn
```

To keep computation time in bounds, we select 100 random time instances, `rbind` the `Spatial` objects, and compute a pooled (pure spatial) variogram:
```{r}
rs = sample(dim(r5to10)[2], 100)
lst = lapply(rs, function(i) { x = r5to10[,i]; x$ti = i; x} )
pts = do.call(rbind, lst)
library(gstat)
v = variogram(PM10~ti, pts[!is.na(pts$PM10),], dX=0)
vmod = fit.variogram(v, vgm(1, "Exp", 200, 1))
plot(v, vmod)
vmod
```

The full, pre-computed space-time variogram is obtained by
```{r eval=FALSE}
rr = rural[,"2005::2010"]
unsel = which(apply(as(rr, "xts"), 2, function(x) all(is.na(x))))
r5to10 = rr[-unsel,]
vv = variogram(PM10~1, r5to10, width=20, cutoff = 200, tlags=0:5)
```

but we will load pre-computed values from package gstat:
```{r}
data(vv, package = "gstat")
vv <- vv[c("np", "dist", "gamma", "id", "timelag", "spacelag")]
```

The following two graphs first show a variogram map in space (x) and time (y), and then a set of spatial variograms where color denotes the time lag:
```{r}
print(plot(vv), split = c(1,1,1,2), more = TRUE)
print(plot(vv, map = FALSE), split = c(1,2,1,2))
```

We see that lag(spatial lag = 0, time lag = 0) is always missing, which is typical for regular data without replicates.

We can fit a metric variogram model to these data by
```{r}
metricVgm <- vgmST("metric",
                   joint=vgm(50,"Exp",100,0),
                   stAni=50)
metricVgm <- fit.StVariogram(vv, metricVgm)
attr(metricVgm, "optim")$value
plot(vv, metricVgm)
```

or a separable model by
```{r}
sepVgm <- vgmST("separable",
                space=vgm(0.9,"Exp", 123, 0.1),
                time =vgm(0.9,"Exp", 2.9, 0.1),
                sill=100)
sepVgm <- fit.StVariogram(vv, sepVgm, method = "L-BFGS-B",
                          lower = c(10,0,0.01,0,1),
                          upper = c(500,1,20,1,200))
attr(sepVgm, "optim")$value
plot(vv, list(sepVgm, metricVgm))
```

These models can then be used for spatiotemporal kriging; the
interested reader is refered to vignettes of package `gstat`, and
examples and demo scripts using the function `krigeST` found there.

### Analysis of events (ST point pattern data)

we'll analyse the events from the foot-and-mouth (fmd) disease data that come with the [stpp](https://cran.uni-muenster.de/package=stpp) package:
```{r}
library(stpp)
data("fmd")
data("northcumbria")
head(fmd)
head(northcumbria)
```
To make the northcumbria polygon more useful, we'll convert it into a `SpatialPolygons` object by
```{r}
nc = rbind(northcumbria, northcumbria[1,]) # close polygon
library(sp)
nc = SpatialPolygons(list(Polygons(list(Polygon(nc)), "NC")))
```
Next, we'll creat a `SpatialPoints` object for the locations
```{r}
pts = SpatialPoints(fmd[,1:2])     # CRS unknown!
plot(nc, axes = TRUE)
plot(pts, cex = 0.5, add = TRUE, col = "#88888888")
```

We can do some simple cell counts by creating a grid first
```{r}
grd = SpatialPixels(SpatialPoints(makegrid(nc, n = 100)))
plot(grd)
plot(nc, add = TRUE)
```

and then assigning a dummy (1) attribute, and aggregating this using `sum`:
```{r}
pts$id = rep(1, length(pts))
image(aggregate(pts, grd, sum))
plot(nc, add = TRUE)
points(pts, cex = .3, col = "#88888888")
title("cell counts")
```

and get an idea of the temporal development by computing average event *time* per cell
```{r}
day = as.Date("2001-01-01") + fmd[,3] # also unknown!
pts$day = day
image(aggregate(pts["day"], grd, mean))
plot(nc, add = TRUE)
points(pts, cex = .3, col = "#88888888")
title("cell mean event time")
```

We can create an irregular space-time object (`STI`) from this, without attributes (DF), by
```{r}
library(spacetime)
sti = STI(pts, day)
plot(sti) # space-time layout
```

This plot shows the space-time (time=x, space=y) layout of the
individual events, where points are simply numbered.  Note that `STI` objects
are *always* time ordered.

We can create a plot that shows attributes if we add an attribute; here we add
a simple index, normalized to $[0,1]$, to indicate temporal distribution.
```{r}
stidf = STIDF(pts, day, data.frame(one = (0:(length(pts)-1))/(length(pts)-1)))
stplot(stidf, sp.layout = nc, key.space = "right", cex = .5,
	xlim = bbox(nc)[1,], ylim = bbox(nc)[2,],
	main = "colour indicates quintile")
```

A variety of the number of observations per grid cell per time interval
is obtained by aggregating `stidf` by a regular space-time grid; we create
this grid by
```{r}
grd = SpatialPixels(SpatialPoints(makegrid(nc, n = 25)))
n = 6 # number of time classes
tcuts = seq(min(index(sti)), max(index(sti)), length.out=n)
grd = STF(grd, tcuts[1:(n-1)])
plot(grd)
```

`STF` creates end times of the time intervals by default by
```{r}
Sys.setenv(TZ = "UTC")
delta(tcuts[1:(n-1)])
```
The number of events per space-time blocks is then computed by
`aggregate`, and plotted with `stplot`:
```{r}
a = aggregate(stidf, grd, sum)
summary(a)
stplot(a, xlim = bbox(nc)[1,], ylim = bbox(nc)[2,], 
	col.regions=gray((45:1)/51), sp.layout = list(nc, first=FALSE),
	main = "number of observations")
```

### Moving objects

A third type of spatiotemporal data concerns trajectories, describing the paths of objects moving through space.  Data consists of sequences of fixes, time-stamped locations measurements. Methods for data analysis have been developed by very different communities, ranging from the transportation and logistics sector to animal ecology. Problems addressed include

1. modelling movement from sample data
1. home range estimation
1. predicting behaviour of individuals, or arrival times
1. studying group behaviour and interactions

Package [trajectories](https://cran.uni-muenster.de/package=trajectories) provides classes and some methods for handling and analysing such data, building upon the work in packages sp and spacetime.
